Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2014-08-14 paper.tex
General updates and some related works.

@@

\begin{abstract}

@@

Finding the locations or regions of highest vorticity or those with the largest norms of the velocity or other fields of interest is
functionality that enables new insights in the study of fluid dynamics. Furthermore, threshold, top-k queries and similarity search in general
are important in many different disciplines \cite{TODO}. In this paper we introduce an efficient evaluation strategy for threshold
queries over time series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
scalar field data stored in the database, but also performs thresholding of derived fields.

%Why is it hard? (E.g., why do naive approaches fail?)
The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
In order to evaluate such a query for a given time step all of the data for that time step must be read and each data point 
will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.
It is impractical to materialize all possible derived fields due to the large size of the datasets and the limits of available storage. Therefore, threshold 
queries have to be evaluated on-demand from the stored data.
Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, as was
shown by Lopez et al. \cite{Lopez} this might not be sufficient for some interactive applications. Moreover, even if the data are available in the cache the
computation associated with the derived field still needs to be performed for each point on the grid as results of previous computations are not cached.

@@

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
datasets. The virtual laboratories built on these systems often do not support
all of the functionality that scientists are interested in. In this paper we present a method for the efficient evaluation of 
threshold queries over numerical simulation data stored in a cluster of relational databases.

%What are the key components of my approach and results? Also include any specific limitations.
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations and moreover introduce an application-aware
cache for query results. We build on the idea of an application-aware cache introduced in \cite{Lopez}, but rather than just data objects we cache
the actual results of the computation, which leads to query performance improvement of up to X times.

@@

\section{JHTDB}
The Johns Hopkins Turbulence Databases (JHTDB), built on top of the GrayWulf and DataScope clusters, serve as a public virtual laboratory for
the study of turbulent phenomena. The JHTDB store several datasets, which are the output of high-resolution numerical simulations of turbulence.
The 3d time-series data are partitioned into small sub-cubes and stored in relational databases distributed across the nodes of the cluster.
Access to the data is provided by means of Web-services and a variety of analysis functions have been
implemented and can be executed through Web-service calls.
At present the service hosts four datasets, which are available publicly. 
The data are the output of numerical simulations of forced isotropic turbulence, magnetohydrodynamics (MHD), channel 
flow turbulence and homogenous buoyancy driven turbulence.
The total amount of space occupied by the datasets is over 230 TB.

The database nodes are part of the GrayWulf \cite{Szalay} and DataScope \cite{TODO} clusters. Each node
is running Windows Server 2008 and SQL Server 2008 R2. The data for each dataset are partitioned spatially across 4 to 8
database nodes, and each database node hosts one or more databases. We use the Morton z-order space-filling
curve to distribute the data across nodes and databases as well as to spatially subdivide each time step
into database atoms, which are of size $8^3$. Each such atom is indexed by the time step, which it belongs to and
by the Morton code of it's lower left corner. This combination of index and data forms a record in the database.
Queries to the data and derived fields, such as derivatives and filtered quantities are evaluated through
stored procedures or user-defined functions implemented in the Common Language Runtime (CLR) framework.

The Web-services are hosted on a front-end Web-server, which handles user requests and hosts the main Webpage portal.
The Web-server acts as a mediator sending the users' requests to the database nodes and initiating their distributed evaluation. 
It assembles the results and sends them back to the client

The JHTDB provide a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
differentiation, particle tracking and spatial filtering. These tasks are often data-intensive and in order to leverage the capabilities of the cluster we
have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. What differentiates threshold and
top-k queries from these tasks is the fact that they usually operate on small subsets of the space or a collection of individual target locations, whereas
threshold and top-k queries usually have to examine the entire data volume. Furthermore, the data product of threshold and top-k queries is much
smaller in relation to the amount of data that need to be examined. This fact combined with the fact that subsequent queries can reuse previously
computed results makes the query results suitable to caching.

@@

The central part of the evaluation strategy for threshold queries that we have developed is the introduction of an application-aware cache
for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Caching the query results
preserves the computational effort in addition to substantially reducing I/O. We do not have to derive the requested field from the raw data for 
queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
a small set of data and do not need to perform any additional computation.

%Application-aware cache that effectively caches not only data, but also computational effort. 
%In the case of thresholding what is cached are the results of the computation. 
%This means that the computation does not have to be repeated and the amount of data cached is several orders of magnitude smaller than the raw data.

The overall execution of threshold queries to the JHTDB is the following. The mediator submits a query to each of the database nodes storing the raw data
asynchronously. Each node begins evaluation of the query by first performing a cache lookup. If the data for the requested field, time-step and spatial region
are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
parameters the query needs to be evaluated from the raw data. In that case the raw data are read into memory with data along the boundary requested from
adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is compared against
the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not have enough space
for the new records, space is freed up by removing the least recently used data. Again, reading from, updating or modifying the cache is done within a
transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction, which
provides for a higher degree of parallelism. Figure \ref{figCacheLookup} illustrates this process.

@@

For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
have to be computed from the stored data. The vorticity is computed from the velocity field by taking its curl. As described in section \ref{science_use_cases}
thresholding the vorticity field is important in the study of fluid dynamics and clustering the locations of maximum vorticity can lead to new insights into
the development of the most intense vortices that are observed. [TODO: Consider showing other fields as well; possibly thresholding the velocity by itself]

@@

-- measure time taken to perform the thresholding computation for an entire time step on a cold cache

@@

\section{Related Work}
A{\ss}falg et al. introduce the concept of threshold queries in time series databases \cite{Asfalg}. Their definition of threshold queries differs from the threshold
queries described in this paper. They are concerned with determining the time series, which exceed a user-defined threshold at time frames similar to
the time series specified in the query. Thus, their definition of threshold queries is concerned with the temporal relationship between the time series
stored in the database (usually one dimensional sequence of measurements) and the time series given in the query. In contrast, our approach focuses on
reporting all of the spatial locations of a multidimensional field where the norm or absolute value of the field exceeds a user prescribed threshold.

>>>>

\newcommand{\kk}[1]{{\color{blue}{\it KK: #1}}}

@@

\begin{abstract}
Large numerical simulations are frequently used to study many different physical phenomena. The datasets produced by these simulations are in the TB and
even PB ranges. Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
The systems built on-top of these architectures, sometimes called virtual laboratories, do not always support the analysis functionality that scientists are 
interested in. In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical simulation datasets
stored in a cluster of relational databases. The method achieves scalability through data-parallel execution. Additionally, we introduce an application-aware 
cache for query results, which has little overhead and improves query performance by X times for queries that hit the cache.

@@

Finding the locations or regions of highest vorticity or those with the largest norms of the velocity or other fields of interest is
functionality that enables new insights in the study of fluid dynamics. Furthermore, threshold, top-$k$ queries and similarity search in general
are important in many different disciplines. In this paper we introduce an efficient evaluation strategy for threshold
queries over time series datasets stored in a cluster of relational databases. Our method evaluates not only threshold queries of the vector or 
scalar field data stored in the database, but also performs thresholding of derived fields.

%Why is it hard? (E.g., why do naive approaches fail?)
The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
In order to evaluate such a query for a given time-step all of the data for that time-step must be read and each data point 
will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.
It is impractical to materialize all possible derived fields due to the large size of the datasets and the limits of available storage. Therefore, threshold 
queries have to be evaluated on-demand from the stored data.
Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, as was
shown by Lopez et al. \cite{Lopez} this might not be sufficient for some interactive applications. Moreover, even if the data are available in the cache the
computation associated with the derived field still needs to be performed for each point on the grid as results of previous computations are not cached.

@@

%Why hasn't it been solved before? (Or, what's wrong with previous proposed solutions? How does mine differ?)
Data-intensive architectures and compute clusters built from commodity hardware rely on parallel I/O to multiple disks and high network bandwidth
to achieve high throughput. Such systems have only recently been deployed for the storage of large numerical simulation 
datasets. The virtual laboratories built on these systems often do not support
all of the functionality that scientists are interested in. Database systems support rollup queries, including top-$k$ queries, but in most cases these
queries apply only simple linear score functions on the attribute values of individual records. 
We present a method for the efficient evaluation of threshold queries over fields derived from the raw vector or scalar fields of the numerical simulation 
stored in the database. Our method computes and operates on derived fields, which are derived by performing kernel computations over the array data 
stored in the database.
Our method makes effective use of the cluster resources and achieves high throughput and scalability.

%What are the key components of my approach and results? Also include any specific limitations.
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations. Each computation is distributed across the nodes of the cluster and
is executed in parallel by several threads within a database node by means of creating multiple connections to each node. Moreover we introduce an 
application-aware cache for query results. We build on the idea of an application-aware cache introduced in \cite{Lopez}, but rather than just data 
objects we cache the actual results of the computation, which leads to query performance improvement of up to X times. In that regard the cache
that we use shares more similarities with the idea of semantic caching \cite{Ren}.

@@

\section{JHTDB}
The method that we have developed for the evaluation of threshold queries of derived fields was deployed and integrated into the 
Johns Hopkins Turbulence Databases (JHTDB). The JHTDB, built on top of the GrayWulf and DataScope clusters, serve as a public virtual laboratory for
the study of turbulent phenomena. The JHTDB store several datasets, which are the output of high-resolution numerical simulations of turbulence.
The 3d time-series data are partitioned into small sub-cubes and stored in relational databases distributed across the nodes of the cluster.
Access to the data is provided by means of Web-services and a variety of analysis functions have been
implemented and can be executed through Web-service calls.
At present the service hosts four datasets, which are available publicly. 
The data are the output of numerical simulations of forced isotropic turbulence, magnetohydrodynamics (MHD), channel 
flow turbulence and homogenous buoyancy driven turbulence.
The total amount of space occupied by the datasets is over 230 TB.

The database nodes are part of the GrayWulf \cite{Szalay} and DataScope \cite{TODO} clusters. Each node
is running Windows Server 2008 and SQL Server 2008 R2. The data for each dataset are partitioned spatially across 4 to 8
database nodes, and each database node hosts one or more databases. We use the Morton z-order space-filling
curve to distribute the data across nodes and databases as well as to spatially subdivide each time-step
into database atoms, which are of size $8^3$. Each such atom is indexed by the time-step, which it belongs to and
by the Morton code of it's lower left corner. This combination of index and data forms a record in the database.
Queries to the data and derived fields, such as derivatives and filtered quantities are evaluated through
stored procedures or user-defined functions implemented in the Common Language Runtime (CLR) framework.

The Web-services are hosted on a front-end Web-server, which handles user requests and hosts the main Web-page portal.
The Web-server acts as a mediator sending the users' requests to the database nodes and initiating their distributed evaluation. 
It assembles the results and sends them back to the client

The JHTDB provide a variety of data-intensive analysis routines that are executed on the database nodes. These include interpolation, 
differentiation, particle tracking and spatial filtering. These tasks are often data-intensive and in order to leverage the capabilities of the cluster we
have developed data-driven batch processing techniques for their evaluation \cite{KanovSC11, KanovSC12}. What differentiates threshold and
top-$k$ queries from these tasks is the fact that they usually operate on small subsets of the space or a collection of individual target locations, whereas
threshold and top-$k$ queries usually have to examine the entire data volume. Furthermore, the data product of threshold and top-$k$ queries is much
smaller in relation to the amount of data that need to be examined. This fact combined with the fact that subsequent queries can reuse previously
computed results makes the query results suitable to caching.

@@

The central part of the evaluation strategy for threshold queries that we have developed is the introduction of an application-aware cache
for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Caching the query results
preserves the computational effort in addition to substantially reducing I/O. We do not have to derive the requested field from the raw data for 
queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
a small set of data and do not need to perform any additional computation.
\kk{Cite results for top-$k$ queries in databases and distributed systems. Remark that non of these strategies operate on large array datasets 
and they do not operate on derived fields, where each value of the derived field depends on multiple values from the original field and requires a kernel
computation accessing multiple data points in a kernel around the target location.}

The overall execution of threshold queries to the JHTDB is the following. The mediator submits a query to each of the database nodes storing the raw data
asynchronously. Each node begins evaluation of the query by first performing a cache lookup. If the data for the requested field, time-step and spatial region
are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
parameters the query needs to be evaluated from the raw data. In that case the raw data are read into memory with data along the boundary requested from
adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is compared against
the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not have enough space
for the new records, space is freed up by removing the least recently used data.  Again, reading from, updating or modifying the cache is done within a
transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel. Figure \ref{figCacheLookup} illustrates this process.

@@

For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
have to be computed from the stored data. The vorticity is computed from the velocity field by taking its curl. As described in section \ref{science_use_cases}
thresholding the vorticity field is important in the study of fluid dynamics and clustering the locations of maximum vorticity can lead to new insights into
the development of the most intense vortices that are observed. \kk{Consider showing other fields as well; possibly thresholding the velocity by itself}

@@

-- measure time taken to perform the thresholding computation for an entire time-step on a cold cache

@@

\section{Related Work}
The processing of top-$k$ queries has been studied extensively in the context of distributed and relational database systems. A survey of different techniques
in the case of centralized processing is given in \cite{Ilyas}. In the case of distributed processing different approaches focus on horizontally 
\cite{Balke, Vlachou} or vertically \cite{Cao, Chaudhuri, Guntzer, Marian, Michel} distributed data. None of these approaches deals with array data stored in
a realtional database system. Moreover, these approaches focus on top-$k$ queries that use simple linear score functions to aggregate the atribute values of
individual records. In addition to top-$k$ queries we are also interested in threshold queries, where the result set includes all locations and values that are
above a user prescribed threshold.

Zhao et al. propose an algorithm for the processing of top-$k$ queries in large-scale distributed environments called BRANCA \cite{Zhao}. They build on the 
idea of semantic caching \cite{Ren} and make use of brach caches, which store results of previous top-$k$ queries with respect to the data stored on each server.
The caching mecahnism that we use is similar in that regard, but the queries that are evaluated in our system operate on derived fields, which are computed
at each location by accessing data from a surrounding region. The queries described in \cite{Zhao} operate over the attributes of individual records only using
simple linear score functions.

A{\ss}falg et al. introduce the concept of threshold queries in time series databases \cite{Asfalg}. Their definition of threshold queries differs from the threshold
queries described in this paper. They are concerned with determining the time series, which exceed a user-defined threshold at time frames similar to
the time series specified in the query. Thus, their definition of threshold queries is concerned with the temporal relationship between the time series
stored in the database (usually one dimensional sequence of measurements) and the time series given in the query. In contrast, our approach focuses on
reporting all of the spatial locations of a multidimensional field where the norm or absolute value of the field exceeds a user prescribed threshold.
